---
title: "Looping"
author: "Brandon Taylor"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Looping}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# Looping

## Introduction
This package contains a flexible framework for extending the pipe into a loop. The basic idea is this: I often run into the problem of wanting to access an unnamed intermediate in a pipe. Why? A basic strategy of working with data frames is to focus on a certain aspect of the data frame, make some changes, and then reincorporate these changes into the original data frame. This work-flow is best understood through illustration.

### Note
This tutorial assumes familiarity with Hadley Wickham's `dplyr` and `magrittr` packages. If you don't know what I'm talking about, go look them up. Your life is about to get a whole lot easier

###Set-up

Import useful libraries for chaining, `knitr` for table output, and of course, `loopr`.
```{r, message=FALSE}
library(loopr)
library(dplyr)
library(magrittr)
library(knitr)
```

Define our loop object.
```{r}
loop = loopClass$new()
```

Set up an extremely simple data frame for illustration.
```{r}
id = c(1, 2, 3, 4)
toFix = c(0, 0, 1, 1)
group = c(1, 1, 1, 0)
example = data_frame(id, toFix, group)
kable(example)
```

## Stack

`loopr` relies on a stack framework. Let's initialize one.

```{r}
stack = stackClass$new()
````

We can `push` data onto the `stack` like this. The names are optional.
```{r}
stack$push(1, name = "first")
stack$push(2, name = "second")
stack$push(3, name = "third")
````

We can `peek` at the top of the `stack`:
```{r}
stack$peek
````

or at the whole thing.

```{r}
stack$stack %>%
  as.data.frame %>%
  kable
````

We can find the `height` of the `stack` as well:
```{r}
stack$height
```

We can also `pop` off items from the `stack`:
```{r}
stack$pop
stack$pop
stack$pop
```

Now the `stack` is empty.
```{r}
stack$stack
```


##Loop
Why is this important? A `loop` object inherits from `stack`.

### Begin
The `begin` method is simply a copy of `push`. After the loop begins, you can focus on any part of your data while still having access to the original data.

```{r}
"first" %>%
loop$begin()
```

### End
To end the loop, you need to merge the data at the beginning of the loop with the data at the end. There are two ending methods defined in `loopr`: `end` and `cross`. Ending the loop takes a function, uses a `pop` from the `loop` `stack` as the first argument to the given function, and its own first argument (or chained argument) as the second.

```{r}
"second" %>%
  loop$end(paste)
```

### Cross
`cross` is nearly identical, but the order of the arguments gets reversed.

```{r}
"first" %>%
  loop$begin()

"second" %>%
  loop$cross(paste)
```

This is much easier to explain in code than in words.

```{r, eval=FALSE}
end(endData, FUN, ...) = FUN(stack$pop, endData, ...)

cross(crossData, FUN, ...) = FUN(crossData, stack$pop, ...)
```

## Ending functions
There are two useful ending functions that are included in this package:`insert` and `amend`. Why are special ending functions needed? In general, traditional join functions are not well suited to the focus-modify-restore work-flow. We need `insert` and `amend` to prioritize information in modified data over information in the original data.

### insert
`insert` is the slightly more simple case. Let's use our example data again.

Create a set of data to `insert`.
```{r}
insertData =
  example %>%
  filter(toFix == 0) %>%
  mutate(toFix = 1) %>%
  select(-group)

kable(insertData)
```

Now let's `insert` it back into the original data.
```{r}
insert(example, insertData, by = "id") %>%
  kable
```

What happened? Where the `by` variables matched, `insert` excised all rows from `example` and inserted `insertData`. At the end, data was sorted by the `by` variable. The `by` variable (or variables) must be included in the function call.

### amend
Let's take a look at the slightly more complicated ending function: `amend`. To understand amend, we first need to understand the underlying column update function.

#### amendColumns
`amendColumns` updates an old set of columns with all non-`NA` values from a matching new set of columns.

Build example data.
```{r}
oldColumn1 = c(0, 0);
newColumn1 = c(1, NA)
oldColumn2 = c(0, 0);
newColumn2 = c(NA, 1)
columnData = data_frame(oldColumn1, newColumn1, oldColumn2, newColumn2)
kable(columnData)
```

Now run `amendColumns`.
```{r}
columnData %>%
  amendColumns(
    c("oldColumn1", "oldColumn2"), 
    c("newColumn1", "newColumn2")) %>%
  kable
```

###fillColumns
There is also a matching function called  fillColumns. In this function, `NA`'s from `newColumn` are replaced with numbers from the `oldColumn`, but nothing else.

```{r}
oldColumn = c(0, 0)
newColumn = c(1, NA)
columnData %>%
  fillColumns(c("newColumn1", "newColumn2"),
              c("oldColumn1", "oldColumn2")) %>%
  kable
```

### amend
`amend` is simply `dplyr::full_join` followed by `amendColumns` to over-write non-key columns from the original dataset with matching-named columns from the new dataset. In this case, `group` from `amendData` overwrites `group` from `example`.

```{r}
amendData = insertData

example %>%
  amend(amendData, by = "id") %>%
  kable
```

If it is not included, `by` defaults to the grouping variables in data.

```{r}
example %>% 
  group_by(id) %>%
  amend(amendData) %>%
  kable
```

A warning: `amend` internally uses the suffix `"toFix"`. If this suffix is already used in your data, modify the `suffix` argument.

##Illustration
Now that we understand how it works, let's use use our `loop`!

Remind ourselves of what the `example` data looks like.

```{r}
kable(example)
```

### Conditional mutation

Here, we convert `toFix` to 0 when `group` is 0.
```{r}
example %>%
  ungroup %>%
  loop$begin() %>%
    filter(group == 0) %>%
    mutate(toFix = 0) %>%
  loop$end(insert, by = "id") %>%
  kable
```

In general, `insert` is best suited to `filter`/`slice` type operations.

### Merged summarize

Here, we summarize toFix in each of the two groups, reverse the results, and then reintegrate the summary into the original data.

```{r}
example %>%
  group_by(group) %>%
  loop$begin() %>%
    summarize(toFix = mean(toFix)) %>%
    mutate(group = rev(group)) %>%
  loop$end(amend) %>%
  kable
```

In general, `amend` is best suited to `summarize`/`do` type operations.

This is only the tip of the iceberg. Do not feel limited to using `amend` and `insert` as ending functions. A whole host of others could be useful: join functions, merge functions, even setNames.

### setNames

Here, we will suffix the names of all the variables within the context of a chain.

```{r}
example %>%
  mutate(group = group + 1) %>%
  loop$begin() %>%
    names %>%
    paste0("Suffix") %>%
  loop$end(setNames) %>%
  kable
```

### bind_rows

Here, we will double the data.

```{r}
example %>%
  mutate(replication = 1) %>%
  loop$begin() %>%
    mutate(replication = 2) %>%
  loop$end(bind_rows) %>%
  kable
```

##Loops within loops
Loops within loops are in fact quite possible. I would be cautious using them. It can be exhilarating, but make sure to indent each loop carefully. Also, it is a good idea to give a name to each loop. This allows one to interpret `loop$stack` for debugging. Here is a quick example that filters the data, replicates the columns, and then re-merges.

```{r}
example %>%
  loop$begin(name = "original") %>%
    filter(group == 1) %>%
    loop$begin(name = "filtered") %>%
       names %>%
       paste0("Extra") %>%
    loop$end(setNames) %>%
    rename(id = idExtra) %>%
  loop$end(amend, by = "id") %>%
  kable
```
